Network Data Model and BerlinMOD Benchmark

### 2010-07-15 Algebras/Network/BerlinMODNetwork/paper/PaperBerlinMODAndNetworkDataModel.tex
added executable SPACE queries

The query set BerlinMOD/R includes 17 queries selected of the set of possible
combinations of the 5 aspects:
\begin{itemize}
  \item known or unknown object identity,
  \item standard attribute, spatial, temporal, or spatio-temporal dimension,
  \item point, range, or unbounded query interval,
  \item single object or object relationships condition type,
  \item with or without aggregation.
\end{itemize}
We will present the 17 queries in more detail in Section
\ref{sec:queries}\todo[caption={check reference}]{Check reference when paper
finished. Parts of query explanations might be moved to the appendix.}
together with our NDM algorithms for these queries.
\section{Implementation of NET}
\label{sec:implNDM}
We started a description of the implementation of NET in \secondo{} in Section
\ref{sec:netdatamod} with the description of the implementation proposed in
\cite{NetworkGueting}. In the meantime this implementation has been changed and
extended in some points to support faster query execution.

@@

The R-Tree indexes of the spatio-temporal bounding boxes of the $trip$ attributes
in $dataMNtrip$ and t$dataSNcar$ relation are different from the
R-Trees of spatio-temporal bounding boxes used in the \bmodb{}.
In the NET representation only the big bounding boxes of the whole trips inserted
in the indexes, whereas in SPACE  representation the much smaller bounding boxes
for each single unit are inserted. This is done because the computation of the
small bounding boxes in NET representation is very expensive (see Section
\ref{sec:implNetDataTyp}). It takes several hours to build the spatio-temporal
unit bounding box indexes for the \dt{mgpoint}  values at the smallest scalefactor.
For bigger scalefactors we stopped the attempt of computation after several days,
because we think that the results of the current used indexes are not so bad that
the additional build time is well invested. We don't expect a great improvement
of query run times.
\subsection{Translate Benchmark Queries}
\label{sec:queries}
\todo[caption={queries appendix}]{Really senseful this way? Move formal queries to Appendix.}
We developed executable \secondo{} queries for each of the 17 BerlinMOD/R queries
for the OBA and the TBA using our network indexes to support faster query execution.
The \secondo{} optimizer is not able to optimize SQL-queries on NDM objects yet, so we tested in
our experiments many different query formulations for each query to get optimal
queries delivering the correct result in a minimum of time.

The limited space does not allow us to show all our executable \secondo{} queries
for the NDM in detail, but they can be downloaded as \secondo{} scripts
from our web page. Here we give only a short overview over the \bmodb{} queries
and their NDM algorithms.

Every time we need a licence in the result or have a query licence number we need
an additional step in the TBA, because we have to join the $dataMNtrip$ and $dataMcar$
relation using the $moid$ attribute and the corresponding B-Tree indexes.
We will not repeat this step at every single TBA query description.

Query 1 asks for the models of the cars with licence plate numbers from $QueryLicences$,
and query 2 for the number of vehicles that are ``passenger cars''. Both queries
deal only with standard attributes; so we only changed the relation names and the
B-Tree indexes to match the NDM representation.

@@

Query 17 asks for the points from $QueryPointsNet$ that have been visited by a
maximum number of different vehicles. In a first step we use almost the query
algorithm from query 4 to select the trips passing a given query
point. After that we group the cars passing query points by the ids of the
query points and count the number of cars passing this query point. In a last
step the point(s) with the maximum number of passing cars is(are) selected.
In the TBA we have to remove duplicate vehicles from the result list before we
count the number of passing cars.
\section{Experimental Setup}
\label{sec:scenario}
\todo[caption={NET SPACE}]{Controll this and next section for right use of NET and SPACE instaed of old abbreviations.}
For our experiments we used a standard personal computer with an AMD Phenom II X4
Quad Core 2.95 GHz CPU, 8 GB main memory, and 2 TB hard disk. We installed the
Linux openSUSE 11.2 as operating system, \secondo{} DBMS version 3.0, and
the \bmodb{} version provided in the web.
\todo[caption={Website}]{Provide scripts in web and add link to scripts!}

We generated three databases with different amounts of data using the data generation
script of the \bmodb{} with the $scalefactor$ 0.05, 0.2, and 1.0. The following
steps are done with all three databases. We first created the \bmodb{} data and indexes
using the script ``BerlinMOD\_CreateObjects.SEC'' for the DMFS. The NDM
representation of the databases was generated by the the script ``Network\_CreateObjects.SEC''
that uses the algorithms and builds the indexes described in Section \ref{sec:Translation}.

Table \ref{tab:dbstat} shows the created amounts of data for the different $scalefactor$
values in both data models. As you can see, the NDM needs less
than 40\% of the storage space of the \bmodb{} data model. The main cause is that
the same trip is represented by less than 50\% of the units in the NDM
compared to the DMFS. This is a very good result and we expect this effect to
increase if the cars make long distance
trips instead of moving in a single town like they do in the benchmark. In towns
cars more often change the street or the velocity than cars that do long distance
trips and so the compact route representation in the NDM should become
more effective than in the town.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabularx}{1.0\textwidth}{|X|c|c|c|c|c|c|}
\hline
&\multicolumn{2}{c|}{\textbf{Scalefactor 0.05}}&\multicolumn{2}{c|}{\textbf{Scalefactor 0.2}}&\multicolumn{2}{c|}{\textbf{Scalefactor 1.0}}\\
\hline
Number of Cars&\multicolumn{2}{c|}{447}&\multicolumn{2}{c|}{894}&\multicolumn{2}{c|}{2000}\\
\hline
Number of Days&\multicolumn{2}{c|}{6}&\multicolumn{2}{c|}{13}&\multicolumn{2}{c|}{28}\\
\hline
Data Generation&\multicolumn{2}{c|}{164.761s}&\multicolumn{2}{c|}{587.299s}&\multicolumn{2}{c|}{3177.46s}\\
\hline
&DFMS&NDM&DFMS&NDM&DFMS&NDM\\
\hline
Data Translation
and Index Build&301.72s&535.65s&1,362.72s&2,190.45s&7,419.13s&11,144.13s\\
\hline
Number of Units&2,646,026&1,260,888&11,296,682&5,346,971&52,140,685&24,697,709\\
\hline
Total Storage Space&2.26 GB&0.86 GB&9.51 GB&3.69 GB&45,76 GB&17.28 GB\\
Data&0.79 GB&0.44 GB&3.35 GB&1.83 GB&15.47 GB& 8.40 GB\\
Indexes&1.48 GB&0.42 GB&6.16 GB&1.86 GB&30.30 GB&8.89 GB\\
\hline
\end{tabularx}
\end{scriptsize}
\caption{Database Statistics}
\label{tab:dbstat}
\end{center}
\end{table}
The long creation time of the NDM representation is
caused by the expensive mapping of spatial and spatio-temporal positions into
network positions. The indexes themselves are built faster in the network representation
than in the \bmodb{} representation because they have less entries and are smaller.

@@

of the \bmodb{} queries and the NDM queries for the OBA and the TBA. We detected

@@

@@

As you can see, the total run time of all queries in the NDM is around
50\% less than the total query run time of the DMFS at each scale factor.

@@

        &\multicolumn{2}{c|}{\textbf{DMFS}}&\multicolumn{2}{c|}{\textbf{NDM}}\\

@@

        \textbf{1}&0.125&0.109&0.128&0.088\\
        \textbf{2}&0.003&0.002&0.003&0.002\\
        \textbf{3}&0.245&0.205&0.227&0.268\\
        \textbf{4}&6.594&7.514&0.238&0.846\\
        \textbf{5}&1.072&1.585&1.098&1.031\\
        \textbf{6}&14.332&6.280&3.995&3.675\\
        \textbf{7}&3.458&3.191&3.893&3.192\\
        \textbf{8}&0.353&0.379&0.201&0.205\\
        \textbf{9}&96.724&166.434&19.840&21.783\\
        \textbf{10}&104.239&31.555&62.972&76.826\\
        \textbf{11}&0.150&0.096&0.224&0.443\\
        \textbf{12}&0.296&0.120&0.202&0.226\\
        \textbf{13}&9.959&6.551&1.094&1.113\\
        \textbf{14}&0.516&0.659&1.566&1.709\\
        \textbf{15}&1.144&0.857&0.579&0.488\\
        \textbf{16}&6.214&14.354&0.612&1.483\\
        \textbf{17}&1.126&0.719&0.228&0.262\\
        \textbf{Total}&246.551&240.609&97.061&113.640\\

@@

        \textbf{1}&0.123&0.096&0.157&0.099\\
        \textbf{2}&0.003&0.003&0.008&0.003\\
        \textbf{3}&0.501&0.327&0.424&0.349\\
        \textbf{4}&32.323&39.113&0.500&6.522\\
        \textbf{5}&1.657&2.985&2.394&2.198\\
        \textbf{6}&57.453&45.931&15.508&13.451\\
        \textbf{7}&17.184&11.150&27.084&22.814\\
        \textbf{8}&0.390&0.379&0.205&0.220\\
        \textbf{9}&250.626&386.452&36.323&46.187\\
        \textbf{10}&447.679&126.870&286.779&287.101\\
        \textbf{11}&0.247&0.165&2.119&2.958\\
        \textbf{12}&4.171&0.182&0.262&0.316\\
        \textbf{13}&27.028&12.106&5.865&4.669\\
        \textbf{14}&1.125&1.143&8.783&9.101\\
        \textbf{15}&7.800&3.874&2.543&1.876\\
        \textbf{16}&6.441&25.298&0.362&0.776\\
        \textbf{17}&8.722&4.042&0.289&0.641\\
        \textbf{Total}&863.472&660.217&389,605&399.281\\

@@

        \textbf{1}&0.196&0.186&0.387&0.185\\
        \textbf{2}&0.005&0.004&0.006&0.004\\
        \textbf{3}&0.731&0.483&1.020&1.349\\
        \textbf{4}&150.172&157.629&2.089&31.769\\
         \textbf{5}&3.274&6.079&5.230&5.494\\
        \textbf{6}&826.483&2002.002&270.468&235.594\\
        \textbf{7}&99.086&53.099&118.840&125.206\\
        \textbf{8}&0.794&0.524&0.254&0.398\\
        \textbf{9}&775.458&2263.531&106.910&143.150\\
        \textbf{10}&3314.518&1942.155&2150.250&1645.812\\
        \textbf{11}&0.685&0.474&6.080&7.889\\
        \textbf{12}&37.445&0.200&0.272&0.290\\
        \textbf{13}&111.587&72.907&26.880&32.540\\
        \textbf{14}&11.397&4.238&36.728&37.700\\
        \textbf{15}&28.512&16.862&9.696&8.602\\
        \textbf{16}&9.726&53.011&0.571&1.880\\
        \textbf{17}&86.357&152.542&0.530&6.884\\
        \textbf{Total}&5456.427&6725.924&2736.210&2284.747\\

@@

For the queries 1 and 2, the query run times are almost the same for all data models
and approaches at the different scale factors. This is what we expected because
both queries deal only with standard attributes and standard indexes, which
are not influenced by the different data models.

For query 3, the run times for all data models and approaches are very small. But
we can see a development of the ratio of the run times between the different
data amounts, data models and approaches. Although the query algorithms for both
data models and approaches are almost the same, the run time ratio for the different
amounts of data is different. For the small databases (scale factor 0.05 and 0.2)
the NDM outperforms in the OBA the DMFS, while for scalefactor 1.0 and all TBA
queries the DMFS outperforms the NDM. We think that
two different effects take place. On the one hand, the number of units in the NDM
is less than the number of units in the DMFS,
such that the unit which contains the query time instant can be found faster.
On the other hand, a \dt{gpoint} value has more internal elements (3 \dt{int}, 1 \dt{real},
and 1 \dt{bool}) than a \dt{point} value (2 \dt{real}, and 1 \dt{bool}), such
that result computing and copying is a little more expensive in the NDM.
The advantage of the smaller number of units in a binary search is smaller if
the number of units becomes bigger and the disadvantage of bigger results becomes
greater; that explains the different run time ratios for query 3.

In query 4 the NDM outperforms the DMFS
significantly at all scale factors ($>$ 2 min OBA, $>$ 6 min TBA at scale factor 1.0).
The NDM index used in query 4 is much smaller (OBA 24 MB, TBA 160 MB, at scalefactor 1.0)
than the spatial unit index of the \bmodb{} (OBA 3.7 GB, TBA 3.7 GB at scalefactor 1.0)
and more precise, such that we do not need an additional refinement step after the
index usage in the NDM, like we do in the DMFS.

We expected the NDM to be slower than the DMFS in the queries 5,6, and 10, because
we retranslate intermediate results from the NDM representation into the DMFS representation.
For query 5 this holds in the OBA. We need a little more time in the NDM
than in the DMFS. But in TBA the NDM outperforms the DMFS. This is due to the fact
that a \dt{gline} value has less \dt{RouteInterval}s
than a \dt{line} value representing the same part of a curve has \dt{HalfSegment}s,
such that the union of two or more \dt{gline} values in the aggregate step of query 5 in
TBA can be computed much faster than the union of two or more \dt{line} values.

The NDM outperforms the DMFS again significantly at query 6
for all amounts of data and approaches. In the NDM we reduce the number of candidate
pairs for the distance computation by pre selecting intersecting extended bounding boxes
and use the operation \op{notEverNearerThan} in OBA and TBA, while in the DMFS
in the OBA no filtering is used and the computation is done by
\op{minimum}(\op{distance}($mp1$,$mp2$)$\leq$10.0), and in the TBA the operation
\op{spatialjoin} is used instead of bounding box intersection. While the operation
\op{distance} has always a run time O($n$) the operation \op{everNearerThan}
stops computation immediately if the distance between two units is less than the
query value to reduce computation time. And the operation \op{spatialjoin} of the
\secondo{} DBMS seems to have a big weakness in implementation. Otherwise the
difference in the TBA query run times could not be so big ($>$ 17 min OBA, $>$
80 min TBA, at scale factor 1.0).\todo{check analysis for correctness.}

After the very good results from query 4 we did not expect query 7 to have such
results in the run times comparison. In fact at scalefactor 1.0 the DMFS outperforms
the NDM significantly in both
approaches and at all scale factors in TBA, while at the smaller scale factors in
OBA the NDM outperforms the DMFS. We think there are two main causes: On the one hand we
have to do the expensive operation \dt{at} for \dt{mgpoint} for the double number
of query \dt{gpoint} compared with the DMFS and
on the other hand the test points out a weakness of the NDM
implementation of the operation \op{at} for \dt{mgpoint}. But in the end, NDM looses
at scale factor 1.0 less than 45 seconds in the OBA respectively less
than 120 seconds in TBA, what is not much compared with the advantages in the
other benchmark queries.

@@

\section{Executable \secondo{} Queries}

@@

the query, and it is a query for the object based approach (OBA), or for the

@@

>>>>

The query set BerlinMOD/R includes 17 queries selected of the set of possible
combinations of the 5 aspects:
\begin{itemize}
  \item known or unknown object identity,
  \item standard attribute, spatial, temporal, or spatio-temporal dimension,
  \item point, range, or unbounded query interval,
  \item single object or object relationships condition type,
  \item with or without aggregation.
\end{itemize}
We will present the 17 queries in more detail in Section
\ref{sec:queries}
together with our NET algorithms for these queries.
\section{Implementation of NET}
\label{sec:implNDM}
We started a description of the implementation of NET in \secondo{} in Section
\ref{sec:netdatamod} with the description of the implementation proposed in
\cite{NetworkGueting}. In the meantime this implementation has been changed and
extended in some points to support faster query execution.

@@

The R-Tree indexes of the spatio-temporal bounding boxes of the $trip$ attributes
in $dataMNtrip$ and t$dataSNcar$ relation are different from the
R-Trees of spatio-temporal bounding boxes used in the \bmodb{}.
In the NET representation only the big bounding boxes of the whole trips inserted
in the indexes, whereas in SPACE  representation the much smaller bounding boxes
for each single unit are inserted. This is done because the computation of the
small bounding boxes in NET representation is very expensive (see Section
\ref{sec:implNetDataTyp}). It takes several hours to build the spatio-temporal
unit bounding box indexes for the \dt{mgpoint}  values at the smallest scalefactor.
For bigger scalefactors we stopped the attempt of computation after several days,
because we think that the results of the current used indexes are not so bad that
the additional build time is well invested. We don't expect a great improvement
of query run times.
\subsection{Translate Benchmark Queries}
\label{sec:queries}
The \secondo{} optimizer is not able to optimize SQL-queries on NET objects yet,
so we tested in our experiments many different query formulations for each of the
17 BerlinMOD/R queries in for the object based approach (OBA) and the trip based
approach (TBA) to get optimal queries delivering the correct result in a minimum of time.

In Appendix \ref{app:executableQueries} you can find the resulting executable
\secondo{} queries for NET in detail. The \secondo{} scripts with the queries
can be downloaded from our web page \cite{NetworkWeb}. In this Section we give
only a short overview over the \bmodb{} queries and their NET algorithms.

Every time we need a licence in the result or have a query licence number we need
an additional step in the TBA queries to connect the $dataMNtrip$ and $dataMcar$
relation using the $moid$ attribute and the corresponding B-Tree indexes.
We will not repeat this step at every single TBA query description.

Query 1 asks for the models of the cars with licence plate numbers from $QueryLicences$,
and query 2 for the number of vehicles that are ``passenger cars''. Both queries
deal only with standard attributes; so we only changed the relation names and the
B-Tree indexes to match the NET representation.

@@

Query 17 asks for the points from $QueryPointsNet$ that have been visited by a
maximum number of different vehicles. In a first step we use almost the query
algorithm from query 4 to select the trips passing a given query
point. After that we group the cars passing query points by the ids of the
query points and count the number of cars passing this query point. In a last
step the point(s) with the maximum number of passing cars is(are) selected.
In the TBA we have to remove duplicate vehicles from the result list before we
count the number of passing cars.
\section{Experimental Setup}
\label{sec:scenario}
For our experiments we used a standard personal computer with an AMD Phenom II X4
Quad Core 2.95 GHz CPU, 8 GB main memory, and 2 TB hard disk. We installed the
Linux openSUSE 11.2 as operating system, \secondo{} DBMS version 3.0, and
the \bmodb{} version provided in the web \cite{NetworkWeb}.

We generated three databases with different amounts of data using the data generation
script of the \bmodb{} with the $scalefactor$ 0.05, 0.2, and 1.0. The following
steps are done with all three databases. We first created the \bmodb{} data and indexes
using the script ``BerlinMOD\_CreateObjects.SEC'' for SPACE. The NET
representation of the databases was generated by the the script ``Network\_CreateObjects.SEC''
that uses the algorithms and builds the indexes described in Section \ref{sec:Translation}.

Table \ref{tab:dbstat} shows the created amounts of data for the different $scalefactor$
values in both data models. As you can see, NET NDM needs less
than 40\% of the storage space of the \bmodb{} data model. The main cause is that
the same trip is represented by less than 50\% of the units in NET
compared to SPACE. This is a very good result and we expect this effect to
increase if the cars make long distance
trips instead of moving in a single town like they do in the benchmark. In towns
cars more often change the street or the velocity than cars that do long distance
trips and so the compact route representation in NET should become
more effective than in the town.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabularx}{1.0\textwidth}{|X|c|c|c|c|c|c|}
\hline
&\multicolumn{2}{c|}{\textbf{Scalefactor 0.05}}&\multicolumn{2}{c|}{\textbf{Scalefactor 0.2}}&\multicolumn{2}{c|}{\textbf{Scalefactor 1.0}}\\
\hline
Number of Cars&\multicolumn{2}{c|}{447}&\multicolumn{2}{c|}{894}&\multicolumn{2}{c|}{2000}\\
\hline
Number of Days&\multicolumn{2}{c|}{6}&\multicolumn{2}{c|}{13}&\multicolumn{2}{c|}{28}\\
\hline
Data Generation&\multicolumn{2}{c|}{164.761s}&\multicolumn{2}{c|}{587.299s}&\multicolumn{2}{c|}{3177.46s}\\
\hline
&SPACE&NET&SPACE&NET&SPACE&NET\\
\hline
Data Translation
and Index Build&301.72s&535.65s&1,362.72s&2,190.45s&7,419.13s&11,144.13s\\
\hline
Number of Units&2,646,026&1,260,888&11,296,682&5,346,971&52,140,685&24,697,709\\
\hline
Total Storage Space&2.26 GB&0.86 GB&9.51 GB&3.69 GB&45,76 GB&17.28 GB\\
Data&0.79 GB&0.44 GB&3.35 GB&1.83 GB&15.47 GB& 8.40 GB\\
Indexes&1.48 GB&0.42 GB&6.16 GB&1.86 GB&30.30 GB&8.89 GB\\
\hline
\end{tabularx}
\end{scriptsize}
\caption{Database Statistics}
\label{tab:dbstat}
\end{center}
\end{table}
The long creation time of the NET representation is
caused by the expensive mapping of spatial and spatio-temporal positions into
network positions. The indexes themselves are built faster in NET representation
than in SPACE representation because they have less entries.

@@

of the SPACE queries and the NET queries for the OBA and the TBA. We detected

@@

\todo{Result Analysis}

@@

As you can see, the total run time of all queries in NET is around
50\% less than the total query run time of SPACE at each scale factor.

@@

        &\multicolumn{2}{c|}{\textbf{SPACE}}&\multicolumn{2}{c|}{\textbf{NET}}\\

@@

        \textbf{1}&0.132&0.109&0.188&0.127\\
        \textbf{2}&0.005&0.004&0.005&0.004\\
        \textbf{3}&0.334&0.256&0.434&0.578\\
        \textbf{4}&9.284&10.559&0.385&1.703\\
        \textbf{5}&3.027&5.484&4.156&4.051\\
        \textbf{6}&51.784&24.941&12.736&7.385\\
        \textbf{7}&7.961&6.364&11.904&8.407\\
        \textbf{8}&0.868&0.768&0.312&0.312\\
        \textbf{9}&343.766&536.425&77.753&75.874\\
        \textbf{10}&397.725&80.306&219.594&155.306\\
        \textbf{11}&0.199&0.145&0.416&1.060\\
        \textbf{12}&0.701&0.154&0.231&0.286\\
        \textbf{13}&32.374&22.030&3.323&3.023\\
        \textbf{14}&0.868&1.026&3.640&4.416\\
        \textbf{15}&3.433&2.323&1.514&1.115\\
        \textbf{16}&20.920&43.683&1.119&2.366\\
        \textbf{17}&1.851&1.691&0.354&0.418\\
        \textbf{Total}&875.229&736.269&338.064&266.432\\

@@

        \textbf{1}&0.146&0.075&0.188&0.122\\
        \textbf{2}&0.006&0.005&0.007&0.005\\
        \textbf{3}&0.478&0.123&0.601&0.891\\
        \textbf{4}&37.458&10.635&0.639&9.892\\
        \textbf{5}&6.049&11.236&8.825&8.567\\
        \textbf{6}&233.646&193.691&55.446&31.201\\
        \textbf{7}&49.711&23.057&51.448&39.682\\
        \textbf{8}&1.034&0.746&0.357&0.357\\
        \textbf{9}&828.867&983.677&135.162&144.155\\
        \textbf{10}&1759.447&239.881&1050.628&579.827\\
        \textbf{11}&0.237&0.135&2.260&3.066\\
        \textbf{12}&9.700&0.201&0.370&0.394\\
        \textbf{13}&78.121&39.199&18.237&12.857\\
        \textbf{14}&1.967&1.949&9.637&10.089\\
        \textbf{15}&23.552&12.614&6.967&10.089\\
        \textbf{16}&21.896&67.620&0.732&1.268\\
        \textbf{17}&14.407&7.430&0.624&1.328\\
        \textbf{Total}&3066.722&1592.275&1342,126&848.655\\

@@

        \textbf{1}&0.218&0.188&0.221&0.183\\
        \textbf{2}&0.010&0.008&0.011&0.008\\
        \textbf{3}&0.756&0.589&0.879&1.345\\
        \textbf{4}&189.155&169.891&1.187&36.939\\
         \textbf{5}&12.248&24.879&20.271&20.693\\
        \textbf{6}&4803.082&8457.233&1010.498&659.709\\
        \textbf{7}&273.725&96.753&222.144&153.983\\
        \textbf{8}&1.968&0.979&0.357&0.424\\
        \textbf{9}&2872.716&3748.291&365.211&424.425\\
        \textbf{10}&13298.981&3020.253&8124.542&3201.532\\
        \textbf{11}&0.755&0.609&6.380&8.454\\
        \textbf{12}&75.108&0.277&0.400&0.438\\
        \textbf{13}&257.358&127.828&94.260&62.879\\
        \textbf{14}&12.598&5.287&38.188&39.453\\
        \textbf{15}&73.552&34.361&18.140&13.329\\
        \textbf{16}&32.352&138.421&0.983&2.735\\
        \textbf{17}&116.552&161.355&1.294&10.084\\
        \textbf{Total}&22021.132&15987.202&9904.964&4636.614\\

@@

For the queries 1 and 2, the query run times are almost the same for all data models
and approaches at the different $scalefactors$. These results are expected, because
both queries deal only with standard attributes and standard indexes, which
are not influenced by the different data models.

Although the query algorithms for both data models and approaches are almost the same,
SPACE outperforms NET for query 3 by tenths of a second. On the one hand, the
number of units in NET is less than the number of units in the SPACE,
such that the unit which contains the query time instant should be found faster.
But on the other hand, a \dt{gpoint} value has more internal elements (3 \dt{int},
1 \dt{real}, and 1 \dt{bool}) than a \dt{point} value (2 \dt{real}, and 1 \dt{bool}),
such that each unit in NET is much bigger than in SPACE, therefore computing and
copying the result needs a little more time in NET than in SPACE.

In query 4 NET outperforms SPACE significantly at all $scalefactors$ ($>$ 3 min OBA,
$>$ 2 min TBA at $scalefactor$ 1.0).
The NET index used in query 4 is much smaller (OBA 24 MB, TBA 160 MB, at $scalefactor$ 1.0)
than the spatial unit index of SPACE (OBA 3.7 GB, TBA 3.7 GB at $scalefactor$ 1.0)
and more precise, such that we do not need an additional expensive refinement step after the
index usage in NET, like we do in SPACE.

We expected NET to be slower than SPACE in the queries 5, 6, and 10, because
we retranslate intermediate results from NET representation into SPACE representation.
For query 5 this holds in the OBA. We need a little more time in NET
than in SPACE. But in TBA NET outperforms SPACE. This is due to the fact
that a \dt{gline} value has less $route\ intervals$
than a \dt{line} value has line segments,
such that the union of \dt{gline} values in the aggregate step of query 5 in
TBA can be computed much faster in NET than the union of \dt{line} values in SPACE.

NET outperforms SPACE again significantly at query 6
for all amounts of data and approaches ($>$ 1 hour for OBA, $>$
2 hours for TBA, at $scalefactor$ 1.0). In NET we reduce the number of candidate
pairs for the distance computation by some filter steps, while in SPACE
in the OBA no filtering is used. NET uses the operation \op{everNearerThan}, which
stops computation immediately if the distance between two units is less than the
query value. Whereas the operation \op{distance} used by SPACE in OBA always checks
all units of the query objects. For the TBA SPACE uses \op{everNearerThan}
instead of \op{distance} and a filter step before. But, whereas NET uses in TBA
a spatio-temporal filtering, SPACE uses only a spatial filtering, such that many
false candidates are included in the expensive distance computation part of the
algorithm.

After the very good results from query 4 we did not expect query 7 to have such
bad results in the run time comparison. In fact SPACE outperforms NET significantly
in both approaches and at all $scalefactors$. in TBA, while at the smaller scale factors in
OBA the NDM outperforms the DMFS. We think there are two main causes: On the one hand we
have to do the expensive operation \dt{at} for \dt{mgpoint} for the double number
of query \dt{gpoint} compared with the DMFS and
on the other hand the test points out a weakness of the NDM
implementation of the operation \op{at} for \dt{mgpoint}. But in the end, NDM looses
at scale factor 1.0 less than 45 seconds in the OBA respectively less
than 120 seconds in TBA, what is not much compared with the advantages in the
other benchmark queries.

@@

\section{Executable \secondo{} NET Queries}
\label{app:executableQueries}

@@

the query, and if it is a query for the object based approach (OBA), or for the

@@

\section{Executable \secondo{} SPACE Queries}
In the sequel we present the executable \secondo{} queries delivered with the
\bmodb{} for the SPACE representation in our experiments.
The name of the query result object indicates the number of
the query, and if it is a query for the object based approach (OBA), or for the
\op{let} OBACRres001 = \\
\>QueryLicences \op{feed} \{O\}\\
\>\op{loopjoin} [ dataSCcar\_Licence\_btree dataSCcar \op{exactmatch} [.Licence\_O]]\\
\>\op{project}[Licence, Model]\\
\op{let} TBACRres001 = \\
\>QueryLicences \op{feed} \{O\}\\
\>\op{loopjoin} [dataMCcar\_Licence\_btree dataMCcar \op{exactmatch}[.Licence\_O]]\\
\>\op{project} [Licence, Model]\\
\op{consume};
\op{let} OBACRres002 = dataSCcar \op{feed filter} [.Type = 'passenger'] \op{count};\\
\op{let} TBACRres002 = dataMCcar \op{feed filter} [.Type = 'passenger'] \op{count};\\
\op{let} OBACRres003 =\\
\>QueryLicences \op{feed head} [10] \{LL\}\\
\>\>\op{loopjoin}[dataSCcar\_Licence\_btree dataSCcar \op{exactmatch}[.Licence\_LL]]\\
\>QueryInstants \op{feed head} [10] \{II\}\\
\>\op{projectextend}[; Licence: .Licence\_LL, Instant: .Instant\_II,\\
\>\>\>\> Pos: \op{val}(.Journey \op{atinstant} .Instant\_II)]\\
\op{let} TBACRres003 =\\
\>QueryLicences \op{feed head} [10] \{LL\}\\
\>\>\op{loopsel} [dataMCcar\_Licence\_btree dataMCcar \op{exactmatch}[.Licence\_LL]\\
\>\>\>\op{project} [Licence, Moid] {LL}]\\
\>\>\>\op{loopjoin} [dataMCtrip\_Moid\_btree dataMCtrip \op{exactmatch} [.Moid\_LL]]\\
\>QueryInstants \op{feed head} [10] \{II\}\\
\>\op{symmjoin}[.Trip \op{present} ..Instant\_II]\\
\>\op{projectextend} [; Licence: .Licence\_LL, Instant: .Instant\_II,\\
\>\>\>\>Pos: \op{val}(.Trip \op{atinstant} .Instant\_II)]\\
\>\op{sortby}[Instant \op{asc}, Licence \op{asc}]\\
\op{let} OBACRres004 =\\
\>QueryPoints \op{feed}\\
\>\op{loopjoin}[ dataSCcar\_Journey\_sptuni \op{windowintersectsS}[\op{bbox}(.Pos)]\\
\>\>\op{sort rdup} dataSCcar \op{gettuples}]\\
\>\op{filter} [.Journey \op{passes} .Pos] \op{project} [Id, Licence]\\
\op{let} TBACRres004 = QueryPoints \op{feed}\\
\>\op{loopjoin}[ dataMCtrip\_Trip\_sptuni \op{windowintersectsS}[\op{bbox}(.Pos)]\\
\>\>\op{sort rdup} dataMCtrip \op{gettuples}]\\
\>\op{filter}[.Trip \op{passes} .Pos]\\
\>\op{project}[Id, Moid]\\
\>\op{loopsel}[ \op{fun}(t1: TUPLE) dataMCcar\_Moid\_btree dataMCcar \op{exactmatch}[attr(t1,Moid)]\\
\>\>\op{projectextend}[Moid; Id: attr(t1,Id), Licence: .Licence]]\\
\>\op{sortby}[Id \op{asc}, Licence \op{asc}]\\
\op{let} OBACRres005tmp1 =\\
\>QueryLicences \op{feed head} [10]\\
\>\op{loopsel}[ dataSCcar\_Licence\_btree dataSCcar \op{exactmatch}[.Licence]]\\
\>\op{projectextend}[Licence; Traj: \op{simplify}(\op{trajectory}(.Journey),0.000001)]\\
\op{let} OBACRres005tmp2 =\\
\>QueryLicences \op{feed head} [20] \op{filter}[.Id $>$ 10]\\
\>\op{loopsel}[ dataSCcar\_Licence\_btree dataSCcar \op{exactmatch}[.Licence]]\\
\>\op{projectextend}[Licence; Traj: \op{simplify}(\op{trajectory}(.Journey),0.000001)]\\
\op{let} OBACRres005 = \\
\>OBACRres005tmp1 \op{feed} \{V1\}\\
\>OBACRres005tmp2 \op{feed} \{V2\}\\
\>\op{projectextend}[ ; Licence1: .Licence\_V1, Licence2: .Licence\_V2, Dist: \op{distance}(.Traj\_V1, .Traj\_V2)]\\
\>\op{sort rdup}\\
\op{delete} OBACRres005tmp1;\\
\op{delete} OBACRres005tmp2;\\
\op{let} TBACRres005Traj1 =\\
\>QueryLicences \op{feed head} [10]\op{project}[Licence] \{LL1\}\\
\>\op{loopsel}[ \op{fun}(t:TUPLE) dataMCcar\_Licence\_btree dataMCcar \op{exactmatch}[attr(t,Licence\_LL1)] \{CAR\}\\
\>\>\op{loopsel}[dataMCtrip\_Moid\_btree dataMCtrip \op{exactmatch}[.Moid\_CAR]]\\
\>\op{projectextend}[;Traj: \op{simplify}(\op{trajectory}(.Trip),0.000001)]\\
\>\op{aggregateB}[Traj; \op{fun}(L1: line, L2: line)\op{ union\_new}(L1,L2); [const line value ()]]\\
\>\op{feed namedtransformstream}[Traj]\\
\>\op{extend}[Licence: attr(t,Licence\_LL1)]]\\
\op{let} TBACRres005Traj2 =\\
\>QueryLicences \op{feed head} [20] \op{filter}[.Id $>$ 10] \{LL1\}\\
\>\op{loopsel}[ \op{fun}(t:TUPLE) dataMCcar\_Licence\_btree dataMCcar \op{exactmatch}[attr(t,Licence\_LL1)] {CAR}\\
\>\>\op{loopsel}[dataMCtrip\_Moid\_btree dataMCtrip \op{exactmatch}[.Moid\_CAR]]\\
\>\op{projectextend}[;Traj: \op{simplify}(\op{trajectory}(.Trip),0.000001)]\\
\>\op{aggregateB}[Traj; \op{fun}(L1: line, L2: line) \op{ union\_new}(L1,L2); [const line value ()]]\\
\>\op{feed namedtransformstream}[Traj]\\
\>\op{extend}[Licence: attr(t,Licence\_LL1)]]\\
\op{let} TBACRres005 = \\
\>TBACRres005Traj1 \op{feed} \{LL1\}\\
\>TBACRres005Traj2 \op{feed} \{LL2\}\\
\>\op{projectextend}[ ; Licence1: .Licence\_LL1, Licence2: .Licence\_LL2, Dist: \op{distance}(.Traj\_LL1, .Traj\_LL2)]\\
\op{delete} TBACRres005Traj1;\\
\op{delete} TBACRres005Traj2;\\
\op{let} OBACRres006 =\\
\>dataSCcar \op{feed} \{V1\} \op{filter}[.Type\_V1 = 'truck']\\
\>dataSCcar \op{feed} \{V2\} \op{filter}[.Type\_V2 = 'truck']\\
\>\op{simmjoin}[.Licence\_V1 $<$ ..Licence\_V2]\\
\>\op{filter}[ \op{minimum}(\op{distance}(.Journey\_V1, .Journey\_V2)) $\leq$ 10.0 ]\\
\>\op{projectextend}[ ; Licence1: .Licence\_V1, Licence2: .Licence\_V2 ]\\
\op{let} TBACRres006BBoxMtrip =\\
\>dataMCcar \op{feed filter}[.Type = 'truck']\\
\>\op{project}[Licence, Moid]\\
\>\op{loopsel}[\op{fun}(t: TUPLE) dataMCtrip\_Moid\_btree dataMCtrip \op{exactmatch}[attr(t,Moid)]\\
\>\>\op{projectextend}[Trip, Moid; BBox: \op{bbox}(.Trip), Licence: attr(t,Licence)]\\
